---
title: "`tidyr` lesson"
output: github_document 
---

```{r, include = F}
library(tidyr)
library(dplyr)
```

## What is tidy data?

A large chunk of your time working with data in `r` is going to be spent cleaning and preparing data. *Tidy* data is a standardized way of organizing data; it was proposed by Hadley Wickham and is designed to make you more effective at exploring and analyzing your data. 

Data is *tidy* if it is in a table that follows these rules: 

* Each column is a variable
* Each row is an observation*

In practical terms this largely translates into getting each variable into it's own column. `tidyr` helps you do this, but more broadly, `tidyr` gives you a set of functions to manipulate data frames. 

### Why tidy data?

* structuring data frames in a consistent way will make you faster at extracting the information you are interested in.
* `dplry` and `ggplot2` and other packages in the tidyverse are  designed to work with tidy data. 

* becoming proficient at manipulating data frames is one of the most important r skills you can develop. 

## Example of tidy data

```{r}
#table1 is tidy
table1
```

## Examples of non-tidy data

Non-tidy data typically has one of two problems but it can have both.

* Variables are spread across multiple columns.
* Observations are scattered across multiple rows.

### Variables spread across multiple columns

```{r}
table4a 
```

###Observations scattered across multiple rows

```{r}
table2
```


## Gathering and spreading

`gather()` and `spread()` are functions designed to help you tidy your data by addressing these two common problems.

### `gather()`

`gather()` lets you gather variables that are spread across multiple columns (i.e. make long/tall data frames). 

```{r}
table4a
```

Conceptually, we want to gather up the two variables `1999` and `2000` into a single variable and create a new variable, `year`, to track which year that case comes from (i.e. 1999 vs 2000).

```{r}
#key is a new variable that tracks which year a value comes from

#value is a new variable that will contain all the data that we gather into a siginle column (i.e. the cases)

#verbose version
gather(data = table4a, key = year, value = cases, `1999`, `2000`)

#piped version; note the use of -country to specify columns to gather
table4a %>% 
  gather(year, cases, -country)

```

### `spread()`

`spread()` is the inverse of gather; it lets you spread observations that are scattered across multiple rows (i.e. make wide data frames).  

```{r}
table2
```

Conceptually, we want to create two new columns in our data frame that represent the variables `cases` and `population`, and we want to fill those variables with the appropriate values from the `count` variable.

```{r}
#key = column that contains the variable names that we want to create
#value = column that contain that values to go into those variables 

#verbose version
spread(data = table2, key = type, value = count)

#piped version
table2 %>% 
  spread(type, count)

```


### Challenge problem 1

1. In `table4b`, year is spread across two columns. 
    a) Do you need `gather()` or `spread()` to tidy this data?
    b) Create a tidy data frame from `table4b`.
    c) Once the data is tidy, find out the overall mean for tb cases.
    d) Using the tidy and non-tidy data frames find the mean tb cases for each year. How do the two approaches (non-tidy vs. tidy) compare? 

##Seperating and uniting

###`seperate()`

`seperate()` splits a column into multiple columns. 

```{r}
table3

#verbose version
separate(data = table3, col = rate, 
         into = c("cases", "population"), sep = "/")

#use for sepertate
table3 %>% 
  separate(rate, c("cases", "population"))

#convert = T makes a guess at an appropriate column type
table3 %>% 
  separate(rate, c("cases", "population"), convert = T)


```

###`unite()`

`unite()` is the inverse of `seperate()`; it lets you combine two columns.

```{r}
#unite 
table5

#verbose; note that it drops the original columns and uses _ to combine
unite(data = table5, col = date, century, year)

unite(data = table5, col = date, century, year, sep = "", remove = F)

#pipe version

table5 %>% 
  unite(col = date, century, year, sep = "")


```

### Challenge problem 2

1. Take a look at `table3` and `table5`. Using either `separate()` or `unite()` make `table3` identical to `table5`. Check your work using the `identical()` function. Hint you might need to look at the help pages for either `unite()` or `separate()`, taking a close look at the sep argument.


##Other useful tidyr functions

##`drop_na()`

`drop_na()` lets you drop rows with missing values

```{r}
smiths

smiths %>% drop_na

```


##`complete()`

`complete()` takes a data frame and makes implicitly missing values explicitly missing. 

```{r}
population

population %>% 
  filter(country == "South Sudan")

population_exp_missing <- 
complete(population, country, year)

population_exp_missing %>% 
  filter(country == "South Sudan")

```


### Challenge problem 3 

1. `drop_na()` removes rows that have missing values on any column, but it can be used in a more focused way (i.e. you can drop rows that have missing data specific columns). Using the `who` dataset (from `tidyr`) make a version of that dataset that has no missing values on the columns `new_sp_m014` and `new_sp_m1524`.

## Other resources and what to learn next

* Hadley Wickham's excellent R for Data Science book has a number of great chapters on tidy data. I used his book to help put together this lesson. The whole book is available for free [here](http://r4ds.had.co.nz). 

* Hadley Wickham's journal article on tidy data, [here](http://vita.had.co.nz/papers/tidy-data.pdf). 

* Check out the join functions that are part of `dplyr`. These help you combine data frames. 

* The `nest()` function helps you easily create data frames that have nested lists (i.e. what Jenny talked about last week); these are very powerful when used with `purrr`.

##Solutions to challenge problems

###Challenge problem 1

```{r}
#abc)

table4b %>% gather(year, cases, -country) %>% 
  summarise(mean_cases = mean(cases))

#d)
#tidy
table4b %>% gather(year, cases, -country) %>% 
  group_by(year) %>% 
  summarise(mean_cases = mean(cases))

#non-tidy
mean(table4b$`1999`)
mean(table4b$`2000`)

table4b %>% summarise(mean_cases_1999 = 
                        mean(`1999`), 
                      mean_cases_2000 = 
                        mean(`2000`))


```

###Challenge problem 2

```{r}
table3
table5

new_table3 <- 
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

identical(new_table3, table5)
```

###Challenge problem 3

```{r}
who %>% drop_na(new_sp_m014, new_sp_m1524) 
```
